### 2014-09-12 Pero_Transcriptome.tex
to Paige

As a direct result of intense heat and aridity, deserts are thought to be amongst the most harsh of environments, particularly for it's mammalian inhabitants. Given the osmoregulation can be challenging for these animals, with failure resulting in death, strong selection should be observed on genes related to the maintenance of water and solute balance. One such animal, \textit{Peromyscus eremicus}, is native to the desert regions of the southwest United States, and may live it's entire life without oral fluid intake. As a first step towards understanding the genetics that underlie this phenotype, we present here a characterization of the transcriptome. We assay four tissues (kidney, liver, brain, testes) from a single individual, and supplement this with population level renal transcriptome sequencing from 15 additional animals. We identified a set of transcripts undergoing both purifying and balancing selection based on estimates of Tajima's D. In addition, we used the branch-site test to identify a transcript, Slc2a9 likely related to desert osmoregulation, undergoing enhanced selection in \peer\: relative to a set of related non-desert rodents.  \\

@@

Deserts are widely considered one of Earth's most harsh environments. Animals living in desert environments are forced to endure intense heat and drought, and as a result, species having evolved in these environments are likely to posses specialized mechanisms that may enhance fitness. While living in deserts likely involves a large number of adaptive traits, the ability to osmoregulate -- to maintain the proper water and electrolyte balance -- appears to be paramount \cite{Walsberg:2000uu}. Indeed, the maintenance of water balance in animals is one of the most important physiologic processes for all organisms, whether they be desert inhabitants or not. Most animals are exquisitely sensitive to changes in osmolality, with slight derangement eliciting physiologic compromise.  When the loss of water exceeds dietary intake, dehydration - and in extreme cases, death - can occur. This process suggests that there is strong selection for mechanisms supporting osmoregulation. Understanding these mechanisms will significantly enhance our understanding of the physiologic processes underlying osmoregulation in extreme environments, having implications for studies of human health, conservation, and climate change. \\

@@

To begin to understand how genes may underlie desert adaptation, I collected 16 individuals from a single population \textit{P. eremicus} over a two year time period (2012-2013). These individuals were captured in live traps, then euthanized using isoflurane overdose and decapitation. Immediately post-mortem, the abdominal and pelvic organs were removed, cut in half (in the case of kidney), placed in RNAlater and flash frozen in liquid Nitrogen. Removal of the brain, with similar preservation techniques, followed that. Time from euthanasia to removal of all organs never exceeded five minutes. Samples were transferred to a -80C freezer at a later date. These procedures were approved by the University of California Berkeley Animal Care and Use Committee and follow guidelines established by the American Society of Mammalogy for the use of wild animals in research \cite{Sikes:2011dz}.

@@

The raw sequence reads corresponding to the four tissue types were error corrected using the software bless \cite{Heo:2014cb}, using kmer=25, based on the developers default recommendations. The error corrected sequence reads were adapter and quality trimmed following recommendations from MacManes \cite{MacManes:2014io} and Mbandi \cite{Christoffels:2014bg}. Specifically, adapter sequence contamination was removed, and low quality nucleotides (defined as \textsc{Phred} \textless 2) were removed using the program Trimmomatic version 0.32  \cite{Lohse:2012fg}. Reads from each tissue were assembled using the Trinity version released 17 July 2014 \cite{Haas:2013jq}. We used flags indicating the stranded nature of sequencing reads and set maximum allowable physical distance between read pairs to 999nt. The assembly was conducted on a linux workstation with 64 cores and 512Gb RAM. To filter the raw sequence assembly, I downloaded \textit{Mus musculus} cDNA and ncRNA datasets from Ensembl (\url{ftp://ftp.ensembl.org/pub/release-75/fasta/mus_musculus/}), and the \textit{Peromyscus maniculatus} reference transcriptome from NCBI (\url{ftp://ftp.ncbi.nlm.nih.gov/genomes/Peromyscus_maniculatus_bairdii/RNA/}). I used a blastN procedure (default settings, evalue set to 10$^{-10}$) to identify contigs in the \textit{P. eremicus} dataset that are likely biological in origin. This procedure, when a reference dataset is available, retains more putative transcripts that a strategy employing expression-based filtering (remove if TMP \textless 1, e.g. \cite{MacManes:2012bu}) of the raw assembly. I then concatenated the filtered assemblies from each tissue into a single file, reducing redundancy using the software cd-hit-est \cite{Li:2006hr} using default setting except that sequences were clustered based on 95\% sequence similarity. This multi-fasta file was used for all subsequent analyses including annotation and mapping. \\

@@

The filtered assemblies were annotated using default settings of the blastN algorithm \cite{Camacho:2009fc} against the Ensembl cDNA and ncRNA datasets described above, downloaded on 1 August 2014. Amongst other things, the Ensemble transcript identifiers were used in the analysis of gene ontology, conducted in the PANTHER package \cite{Mi:2004iv}. Next, because rapidly evolving nucleotide sequences may evade detection by blast algorithms, we used HMMER3 \cite{Wheeler:2013gj} to search for conserved protein domains contained in the dataset using the Pfam database \cite{Punta:2012ko}. Lastly, I extracted putative coding sequences using Transdecoder version 4Jul2014 (\url{http://transdecoder.sourceforge.net/})\\

To identify patterns of gene expression unique to each tissue type, I mapped sequence reads from each tissue type to the reference assembly using bwa-mem \cite{Li:2013wn}. We estimated expression individually for the four tissues using default settings of the software eXpress \cite{Roberts:2012dh}. Interesting patterns of expression, including instances where expression was limited to a single tissue type were identified and visualized. \\

@@

I then estimated gene expression on each of these tissue-specific datasets, which allowed me to understand expression patterns in the multiple tissues. Specifically, I constructed a Venn diagram ({\hypertarget{Figure 1}{Figure 1}}), which allowed me to visualize the proportion genes whose expression was limited to a single tissue, and those where expression was ubiquitous. 66324 transcripts are expressed on all tissue types, while 13086 are uniquely expressed in kidney, 2222 in the testes, 3580 in the brain, and 2798 in the liver. The kidney appears to an outlier in the number of unique sequences, though this could be the result of the recovery of more lowly expressed transcripts that may be the result of deeper sequencing.  \\

In addition to this, I estimated mean TMP (number of transcripts per million) for all transcripts. {\hypertarget{Table 2}{Table 2}} consists of the 10 genes whose mean TMP was the highest. Several genes in this list are present predominately in a single tissue type. For instance Transcript\_126459, Albumin is very highly expressed in the liver, but less so in the other tissues. It should be noted, however, that making inference based on uncorrected values for TPM is not warranted. Statistical testing for differential expression was not implemented due to the fact that no replicates are available.  \\

After expression estimation, the filtered assemblies were concatenated together, and after removal of redundancy with cd-hit-est, 123,123 putative transcripts remained (To be available on Genbank). From this filtered concatenated dataset, I extracted 71626 putative coding sequences (72Mb, to be available on Dryad). Of these 71626 sequences, 38221 were complete exons (containing both start and stop codons), while other were either truncated at the 5-prime end (20239 sequences), 3-prime end (6445 sequences), or were internal (6721 sequencing having neither stop nor start codon). The results of a Pfam search conducted on the predicted amino acid sequences will be found on Dryad. \\

@@

\textbf{\hypertarget{Figure 2}{Figure 2}} \\
\centerline{\includegraphics[width=18.0\baselineskip]{/Users/macmanes/Documents/pero_transcriptome/tajD.jpeg}}
\begin{quote}
\small{Figure 2. The distribution of Tajima's D for all putative transcripts.}
\end{quote}  
\vspace{10mm}
\textbf{\hypertarget{Table 3}{Table 3}} \\
\begin{center}
\begin{adjustwidth}{-.75in}{-.75in}% adjust the L and R margins by 1 inch
\begin{tabular}{ l l l l }
\textbf{Transcript ID} & \textbf{GenBank ID} & \textbf{Description} & \textbf{Tajima's D}\\
\hline
Transcript\_49049 & XM\_006533884.1 & heterogeneous nuclear ribonucleoprotein H1 (Hnrnph1) & 3.26\\
Transcript\_38378 & XM\_006522973.1 & Son DNA binding protein (Son) & 3.19\\
Transcript\_126187 & NM\_133739.2 & transmembrane protein 123 (Tmem123) & 3.02\\
Transcript\_70953 & XM\_006539066.1 & chloride channel Kb (Clcnkb) & 2.96 \\
Transcript\_37736 & XM\_006997718.1 & h-2 class I histocompatibility antigen & 2.92 \\
Transcript\_21448 & XM\_006986148.1 & zinc finger protein 624-like & 2.84\\
Transcript\_47450 & NM\_009560.2 zinc & finger protein 60 (Zfp60) & 2.82\\
Transcript\_122250 & XM\_006539068.1 & chloride channel Kb (Clcnkb) & 2.81\\
Transcript\_78367 & XM\_006496814.1 & CDC42 binding protein kinase alpha (Cdc42bpa) & 2.78 \\
Transcript\_96470 & XM\_006987129.1 & interferon-inducible GTPase 1-like & 2.77 \\
 \end{tabular}
\begin{quote}
\small{Table 3. The 10 transcripts with the highest values for Tajima's D, which is suggestive of balancing selection.}
\end{quote}  
\end{adjustwidth}
\end{center}
\vspace{10mm}
\textbf{\hypertarget{Table 4}{Table 4}} \\
\begin{center}
\begin{adjustwidth}{-.75in}{-.75in}% adjust the L and R margins by 1 inch
\begin{tabular}{ l l l l }
\textbf{Transcript ID} & \textbf{GenBank ID} & \textbf{Description} & \textbf{Tajima's D} \\
\hline
Transcript\_84359 & XM\_006991127.1 & nuclear receptor coactivator 3 (Ncoa3) & -2.82\\
Transcript\_87121 & XM\_006970128.1 & methyl-CpG binding domain protein 2 (Mbd2) & -2.82 \\
Transcript\_125755 & EU053203.1 & alpha globin gene cluster & -2.78\\
Transcript\_87128 & XM\_006976644.1 & membrane-associated ring finger (March5) & -2.76 \\
Transcript\_55468 & XM\_006978377.1 & Vpr (HIV-1) binding protein (Vprbp) & -2.75\\
Transcript\_116042 & XM\_006980811.1 & membrane associated guanylate kinase (Magi3) & -2.75  \\
Transcript\_18966 & XM\_006982814.1 & ubiquitin protein ligase E3 component n-recognin 5 (Ubr5) & -2.75 \\
Transcript\_122204 & XM\_008772511.1 & zinc finger protein 612 (Zfp612) & -2.75 \\
Transcript\_100550 & XM\_006971297.1 & bromodomain adjacent to zinc finger domain, 1B (Baz1b) & -2.74\\
Transcript\_33267 & XM\_006975561.1 & pumilio RNA-binding family member 1 (Pum1) & -2.75\\
 \end{tabular}
\begin{quote}
\small{Table 4. The 10 transcripts with the lowest values for Tajima's D, which is suggestive of purifying or directional selection.}
\end{quote}  
\end{adjustwidth}
\end{center}

@@

To begin to test the hypothesis that selection on transcripts related to osmoregulation is enhanced in the desert adapted \peer, I implemented the branch-site test as described above, setting the sequence corresponding to \peer\:for both Slc2a9 and Vdr as the foreground lineages in 2 distinct program executions. These two transcripts were chose specifically because they, the former significantly, were recently linked to osmoregulation in a desert rodent \cite{Marra:2014de}. The test for Slc2a9 was highly significant (2$\Delta$Lnl=51.4, df=1, p=0), indicating enhanced selection in \peer\: relative to the other lineages. The branch site test for positive selection conducted on the Vdr gene was non-significant (2$\Delta$Lnl=0.68, df=1, p=1). This limited analysis of selection is to be followed up by an analysis of genome wide patterns of natural selection. \\

>>>>

As a direct result of intense heat and aridity, deserts are thought to be amongst the most harsh of environments, particularly for their mammalian inhabitants. Given that osmoregulation can be challenging for these animals, with failure resulting in death, strong selection should be observed on genes related to the maintenance of water and solute balance. One such animal, \textit{Peromyscus eremicus}, is native to the desert regions of the southwest United States, and may live its entire life without oral fluid intake. As a first step towards understanding the genetics that underlie this phenotype, we present here a characterization of the \peer\: transcriptome. We assay four tissues (kidney, liver, brain, testes) from a single individual, and supplement this with population level renal transcriptome sequencing from 15 additional animals. We identified a set of transcripts undergoing both purifying and balancing selection based on estimates of Tajima's D. In addition, we used the branch-site test to identify a transcript, Slc2a9 likely related to desert osmoregulation, undergoing enhanced selection in \peer\: relative to a set of related non-desert rodents.  \\

@@

Deserts are widely considered one of Earth's most harsh environments. Animals living in desert environments are forced to endure intense heat and drought, and as a result, species living in these environments are likely to posses specialized mechanisms to deal with them. While living in deserts likely involves a large number of adaptive traits, the ability to osmoregulate -- to maintain the proper water and electrolyte balance -- appears to be paramount \cite{Walsberg:2000uu}. Indeed, the maintenance of water balance in animals is one of the most important physiologic processes for all organisms, whether they be desert inhabitants or not. Most animals are exquisitely sensitive to changes in osmolality, with slight derangement eliciting physiologic compromise.  When the loss of water exceeds dietary intake, dehydration - and in extreme cases, death - can occur. Thus there has likely been strong selection for mechanisms supporting optimal osmoregulation in species that live where water is limiting. Understanding these mechanisms will significantly enhance our understanding of the physiologic processes underlying osmoregulation in extreme environments, having implications for studies of human health, conservation, and climate change. \\

@@

To begin to understand how genes may underlie desert adaptation, we collected 16 individuals from a single population \textit{P. eremicus} over a two year time period (2012-2013). These individuals were captured in live traps, then euthanized using isoflurane overdose and decapitation. Immediately post-mortem, the abdominal and pelvic organs were removed, cut in half (in the case of kidney), placed in RNAlater and flash frozen in liquid Nitrogen. Removal of the brain, with similar preservation techniques, followed that. Time from euthanasia to removal of all organs never exceeded five minutes. Samples were transferred to a -80C freezer at a later date. These procedures were approved by the University of California Berkeley Animal Care and Use Committee and follow guidelines established by the American Society of Mammalogy for the use of wild animals in research \cite{Sikes:2011dz}.

@@

The raw sequence reads corresponding to the four tissue types were error corrected using the software bless \cite{Heo:2014cb}, using kmer=25, based on the developers default recommendations. The error corrected sequence reads were adapter and quality trimmed following recommendations from MacManes \cite{MacManes:2014io} and Mbandi \cite{Christoffels:2014bg}. Specifically, adapter sequence contamination was removed, and low quality nucleotides (defined as \textsc{Phred} \textless 2) were removed using the program Trimmomatic version 0.32  \cite{Lohse:2012fg}. Reads from each tissue were assembled using the Trinity version released 17 July 2014 \cite{Haas:2013jq}. We used flags indicating the stranded nature of sequencing reads and set maximum allowable physical distance between read pairs to 999nt. The assembly was conducted on a linux workstation with 64 cores and 512Gb RAM. To filter the raw sequence assembly, we downloaded \textit{Mus musculus} cDNA and ncRNA datasets from Ensembl (\url{ftp://ftp.ensembl.org/pub/release-75/fasta/mus_musculus/}), and the \textit{Peromyscus maniculatus} reference transcriptome from NCBI (\url{ftp://ftp.ncbi.nlm.nih.gov/genomes/Peromyscus_maniculatus_bairdii/RNA/}). We used a blastN procedure (default settings, evalue set to 10$^{-10}$) to identify contigs in the \textit{P. eremicus} dataset that are likely biological in origin. This procedure, when a reference dataset is available, retains more putative transcripts that a strategy employing expression-based filtering (remove if TMP \textless 1, e.g. \cite{MacManes:2012bu}) of the raw assembly. We then concatenated the filtered assemblies from each tissue into a single file, reducing redundancy using the software cd-hit-est \cite{Li:2006hr} using default setting except that sequences were clustered based on 95\% sequence similarity. This multi-fasta file was used for all subsequent analyses including annotation and mapping. \\

@@

The filtered assemblies were annotated using default settings of the blastN algorithm \cite{Camacho:2009fc} against the Ensembl cDNA and ncRNA datasets described above, downloaded on 1 August 2014. Amongst other things, the Ensemble transcript identifiers were used in the analysis of gene ontology, conducted in the PANTHER package \cite{Mi:2004iv}. Next, because rapidly evolving nucleotide sequences may evade detection by blast algorithms, we used HMMER3 \cite{Wheeler:2013gj} to search for conserved protein domains contained in the dataset using the Pfam database \cite{Punta:2012ko}. Lastly, we extracted putative coding sequences using Transdecoder version 4Jul2014 (\url{http://transdecoder.sourceforge.net/})\\

To identify patterns of gene expression unique to each tissue type, we mapped sequence reads from each tissue type to the reference assembly using bwa-mem \cite{Li:2013wn}. We estimated expression individually for the four tissues using default settings of the software eXpress \cite{Roberts:2012dh}. Interesting patterns of expression, including instances where expression was limited to a single tissue type were identified and visualized. \\

@@

We then estimated gene expression on each of these tissue-specific datasets, which allowed us to understand expression patterns in the multiple tissues. Specifically, we constructed a Venn diagram ({\hypertarget{Figure 1}{Figure 1}}), which allowed us to visualize the proportion genes whose expression was limited to a single tissue, and those where expression was ubiquitous. 66324 transcripts are expressed on all tissue types, while 13086 are uniquely expressed in kidney, 2222 in the testes, 3580 in the brain, and 2798 in the liver. The kidney appears to an outlier in the number of unique sequences, though this could be the result of the recovery of more lowly expressed transcripts that may be the result of deeper sequencing.  \\

In addition to this, we estimated mean TMP (number of transcripts per million) for all transcripts. {\hypertarget{Table 2}{Table 2}} consists of the 10 genes whose mean TMP was the highest. Several genes in this list are present predominately in a single tissue type. For instance Transcript\_126459, Albumin is very highly expressed in the liver, but less so in the other tissues. It should be noted, however, that making inference based on uncorrected values for TPM is not warranted. Statistical testing for differential expression was not implemented due to the fact that no replicates are available.  \\

After expression estimation, the filtered assemblies were concatenated together, and after removal of redundancy with cd-hit-est, 123,123 putative transcripts remained (To be available on Genbank). From this filtered concatenated dataset, we extracted 71626 putative coding sequences (72Mb, to be available on Dryad). Of these 71626 sequences, 38221 were complete exons (containing both start and stop codons), while other were either truncated at the 5-prime end (20239 sequences), 3-prime end (6445 sequences), or were internal (6721 sequencing having neither stop nor start codon). The results of a Pfam search conducted on the predicted amino acid sequences will be found on Dryad. \\

@@

\textbf{\hypertarget{Figure 2}{Figure 2}} \\
\centerline{\includegraphics[width=18.0\baselineskip]{/Users/macmanes/Documents/pero_transcriptome/tajD.jpeg}}
\begin{quote}
\small{Figure 2. The distribution of Tajima's D for all putative transcripts.}
\end{quote}  
\vspace{10mm}
\textbf{\hypertarget{Table 3}{Table 3}} \\
\begin{center}
\begin{adjustwidth}{-.75in}{-.75in}% adjust the L and R margins by 1 inch
\begin{tabular}{ l l l l }
\textbf{Transcript ID} & \textbf{GenBank ID} & \textbf{Description} & \textbf{Tajima's D}\\
\hline
Transcript\_49049 & XM\_006533884.1 & heterogeneous nuclear ribonucleoprotein H1 (Hnrnph1) & 3.26\\
Transcript\_38378 & XM\_006522973.1 & Son DNA binding protein (Son) & 3.19\\
Transcript\_126187 & NM\_133739.2 & transmembrane protein 123 (Tmem123) & 3.02\\
Transcript\_70953 & XM\_006539066.1 & chloride channel Kb (Clcnkb) & 2.96 \\
Transcript\_37736 & XM\_006997718.1 & h-2 class I histocompatibility antigen & 2.92 \\
Transcript\_21448 & XM\_006986148.1 & zinc finger protein 624-like & 2.84\\
Transcript\_47450 & NM\_009560.2 zinc & finger protein 60 (Zfp60) & 2.82\\
Transcript\_122250 & XM\_006539068.1 & chloride channel Kb (Clcnkb) & 2.81\\
Transcript\_78367 & XM\_006496814.1 & CDC42 binding protein kinase alpha (Cdc42bpa) & 2.78 \\
Transcript\_96470 & XM\_006987129.1 & interferon-inducible GTPase 1-like & 2.77 \\
 \end{tabular}
\begin{quote}
\small{Table 3. The 10 transcripts with the highest values for Tajima's D, which is suggestive of balancing selection.}
\end{quote}  
\end{adjustwidth}
\end{center}
\vspace{10mm}
\textbf{\hypertarget{Table 4}{Table 4}} \\
\begin{center}
\begin{adjustwidth}{-.75in}{-.75in}% adjust the L and R margins by 1 inch
\begin{tabular}{ l l l l }
\textbf{Transcript ID} & \textbf{GenBank ID} & \textbf{Description} & \textbf{Tajima's D} \\
\hline
Transcript\_84359 & XM\_006991127.1 & nuclear receptor coactivator 3 (Ncoa3) & -2.82\\
Transcript\_87121 & XM\_006970128.1 & methyl-CpG binding domain protein 2 (Mbd2) & -2.82 \\
Transcript\_125755 & EU053203.1 & alpha globin gene cluster & -2.78\\
Transcript\_87128 & XM\_006976644.1 & membrane-associated ring finger (March5) & -2.76 \\
Transcript\_55468 & XM\_006978377.1 & Vpr binding protein (Vprbp) & -2.75\\
Transcript\_116042 & XM\_006980811.1 & membrane associated guanylate kinase (Magi3) & -2.75  \\
Transcript\_18966 & XM\_006982814.1 & ubiquitin protein ligase E3 component n-recognin 5 (Ubr5) & -2.75 \\
Transcript\_122204 & XM\_008772511.1 & zinc finger protein 612 (Zfp612) & -2.75 \\
Transcript\_100550 & XM\_006971297.1 & bromodomain adjacent to zinc finger domain, 1B (Baz1b) & -2.74\\
Transcript\_33267 & XM\_006975561.1 & pumilio RNA-binding family member 1 (Pum1) & -2.75\\
 \end{tabular}
\begin{quote}
\small{Table 4. The 10 transcripts with the lowest values for Tajima's D, which is suggestive of purifying or directional selection.}
\end{quote}  
\end{adjustwidth}
\end{center}

@@

To begin to test the hypothesis that selection on transcripts related to osmoregulation is enhanced in the desert adapted \peer, we implemented the branch-site test as described above, setting the sequence corresponding to \peer\:for both Slc2a9 and Vdr as the foreground lineages in 2 distinct program executions. These two transcripts were chose specifically because they, the former significantly, were recently linked to osmoregulation in a desert rodent \cite{Marra:2014de}. The test for Slc2a9 was highly significant (2$\Delta$Lnl=51.4, df=1, p=0), indicating enhanced selection in \peer\: relative to the other lineages. The branch site test for positive selection conducted on the Vdr gene was non-significant (2$\Delta$Lnl=0.68, df=1, p=1). This limited analysis of selection is to be followed up by an analysis of genome wide patterns of natural selection. \\
